---
title: "STATS 250 Lab 12"
author: "Lab Dream Team"
date: "Week of 11/16/2020"
output:
  html_document:
  toc: true
toc_float: true
toc_collapsed: true
theme: lumen
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require(checkmate))
  install.packages('checkmate')
source("plotT.R")
```

## Learning Objectives

### Statistical Learning Objectives
1. Continue discussing quantitative data, this week in regards to a paired mean and a difference in two means scenario.
1. Understand whether data is considered paired or from two independent samples.

### R Learning Objectives
1. Create a "difference" variable in a data.frame
1. Learn how to use R to perform paired t-tests and t-tests for two independent samples.


### Functions covered in this lab
1. `pt()`
1. `qt()`
1. `plotT()` (custom)
1. `qqnorm()`
1. `qqline()`
1. `t.test()`

### Weekly Advice
The function `t.test()` will compute the results from both a hypothesis test and a confidence interval. You should pay close attention to the alternative hypothesis. If you are running a hypothesis test, be sure that the correct direction (`greater`, `less`, `two.sided`) is specified in `t.test()`, otherwise your p-value will be inaccurate. If you are running a confidence interval, be sure that the direction is set to `two.sided`, which can be achieved by leaving the alternative hypothesis as the default. You can double-check your thinking by reading the output - the alternative hypothesis is printed in the 4th line of output.
  
## Lab Demo

### Name That Scenario!

Before we get started, let's take a moment to review what we have done so far this semester. We have come across many scenarios, each having a unique set of conditions and computations. To do this, let us fondly recall the delicious chocolate chip cookies at the Mosher-Jordan (MoJo) dining hall.

For each description of a research question below, select the one corresponding statistical analysis technique most appropriate for addressing that research question.

#### Scenario 1

Do you get cookies with your dinner? A UM student decided to stop by MoJo one day at dinner, and asks each student to report the number of cookies they ate while at MoJo. From this data, an appropriate random sample will be selected. The UM student wishes to see if on average, the UM students ate more than one cookie at MoJo during their dinner.

(a) One-sample Z-test for a population proportion
(b) Two-sample Z-test for the comparison of two population proportions
(c) One-sample t-test for a population mean
(d) Paired t-test for a population mean of the difference
(e) Two-sample t-test for the comparison of two population means

The answer is *replace this text with your answer* because *replace this text with your reasoning*.

#### Scenario 2

Does MoJo cure homesickness? A UM student is interested in seeing if perhaps cookies are just a fond memory for freshman students embarking on their first semester away from home. She gathers a large random sample of UM freshmen students, to ask whether or not they enjoy the cookies at MoJo. Next, she gathers a large random sample of UM students who are not freshmen, to ask whether or not they enjoy the cookies at MoJo. She hopes to compare the differences in these two rates.

(a) One-sample Z-test for a population proportion
(b) Two-sample Z-test for the comparison of two population proportions
(c) One-sample t-test for a population mean
(d) Paired t-test for a population mean of the difference
(e) Two-sample t-test for the comparison of two population means

The answer is *replace this text with your answer* because *replace this text with your reasoning*.

#### Scenario 3

Are more cookies made at MoJo than at East Quad, on average? A UM student decides to select a random sample of 30 days from the Winter 2019 semester. For each of those 30 days, they ask each dining hall to report the number of cookies baked. These results will be used to assess whether more cookies are made at MoJo than at East Quad, on average.

(a) One-sample Z-test for a population proportion
(b) Two-sample Z-test for the comparison of two population proportions
(c) One-sample t-test for a population mean
(d) Paired t-test for a population mean of the difference
(e) Two-sample t-test for the comparison of two population means

The answer is *replace this text with your answer* because *replace this text with your reasoning*.

#### Scenario 4

Who has the tastiest cookies? A UM student decides to stop by MoJo to get 40 freshly baked cookies, then stops at East Quad for another 40 freshly baked cookies. She then gets a random sample of 40 UM freshmen, and has them each take a blind taste test. They will taste each cookie, one at a time, without knowing its origin, and select the cookie they like the most. The UM student would like to see if a majority pick MoJo cookies as the tastiest. 

(a) One-sample Z-test for a population proportion
(b) Two-sample Z-test for the comparison of two population proportions
(c) One-sample t-test for a population mean
(d) Paired t-test for a population mean of the difference
(e) Two-sample t-test for the comparison of two population means

The answer is *replace this text with your answer* because *replace this text with your reasoning*.

#### Scenario 5

Do athletes love cookies? As a follow up to a previous observation, a UM student decides that it might be best to gather data about whether the student is an athlete or not, as the number of cookies they eat in a week might differ, on average. The UM student gathers a large random sample of UM athletes to ask them to self-report the number of cookies they ate at MoJo the week before. Then the UM student gathers a large random sample of UM non-athletes to ask them to self-report the number of cookies they ate at MoJo the week before. 

(a) One-sample Z-test for a population proportion
(b) Two-sample Z-test for the comparison of two population proportions
(c) One-sample t-test for a population mean
(d) Paired t-test for a population mean of the difference
(e) Two-sample t-test for the comparison of two population means

The answer is *replace this text with your answer* because *replace this text with your reasoning*.


### Paired Data

To explore paired t-tests in R, we're going to use some data on textbook prices to address the question of whether textbooks purchased online are cheaper than at the UCLA bookstore. The data contains information about the textbooks used in 73 UCLA courses in Spring 2010 (this is less than 10% of all of UCLA's courses that semester).   

```{r textbookData}
textbooks <- read.csv("textbooks.csv")
head(textbooks)
```

Each textbook has two prices: one price for the UCLA bookstore (`ucla_new`) and one price for Amazon. There is a natural correspondence between these prices, since they're for the same book! So we'll pair the data. 

A paired t-test uses the same procedure as a t-test for one population mean, just using an average *difference*. The test statistic is $$t = \frac{\bar{x}_{\text{diff}} - \mu_{0}}{s_{\text{diff}} / \sqrt{n}}.$$

In practice, that means we can use exactly the same R machinery to conduct a paired t-test, just using a variable that represents the differences we're interested. Let's create such a variable that represents the bookstore price minus the Amazon price..

```{r priceDiffs, error = T}
# to create a new variable in the textbooks data.frame, we use the assignment
# operator (<-) like we'd normally use to give something a name. Here, we're
# saying the "diff" variable should go inside of the textbooks data.frame.

textbooks$diff <- ___________ - ___________
head(textbooks)
```

So now that we've got our differences variable, let's just run a t-test. We want to know if there's a *difference* between the prices, on average, so our hypotheses are
$$H_0: \ \mu_{\text{diff}} = 0 \quad \text{vs.} \quad H_{a}: \ \mu_{\text{diff}} \neq 0,$$
where $\mu_{\text{diff}}$ is the population average difference in textbook prices for all courses at UCLA (bookstore minus Amazon). 

Before we can proceed, we should check conditions:

1. Independence: we're told this is less than 10% of the courses at UCLA so we'll proceed assuming that the textbooks in the sample are independent of each other.
2. Nearly normal: Let's look at a histogram and Q-Q plot!

```{r diffConditions, error = T}
hist(textbooks$diff, main = "Histogram of Price Differences",
     xlab = "Price Difference (bookstore minus Amazon)")

qqnorm(textbooks$diff)
qqline(textbooks$diff)
```

*Replace this text with your conclusion about the nearly normal condition. Can we proceed anyway?*

Now we can just run t.test() on our `diff` variable:

```{r pairedTTest, error = T}
t.test(textbooks$diff, mu = 0, alternative = "two.sided")
```

*Replace this text with your conclusion about the population average difference in textbook prices*


### Difference of Two Means

```{r penguins}
penguins <- read.csv("https://raw.githubusercontent.com/STATS250SBI/palmerpenguins/master/inst/extdata/penguins_NArm.csv")
```

Remember this bimodal histogram from last week?

```{r flipperHist}
hist(penguins$flipper_length_mm,
     main = "Histogram of Flipper Length",
     xlab = "Flipper length (mm)",
     col = "peachpuff")
```

Bimodal distributions tend to suggest that there's an important subgroup effect happening. It's pretty clear from our knowledge of this data and the bimodal distribution of flipper length that penguin species is probably related to things here.

In the chunk below, take a few minutes to write some code that will help you investigate whether penguin species is related to flipper length.

```{r investigateSpecies, error = T}

```

We can already sort of tell from this that Gentoo penguins seem to be bigger than Adelie and Chinstrap penguins, so there's probably not much point seeing if their average flipper length is different from one of the other species'. So, let's compare the mean flipper lengths of Adelie and Chinstrap penguins. Let's formulate some hypotheses.

$H_0:$ Replace this text with your null hypothesis.

$H_a:$ Replace this text with your alternative hypothesis.

Before we proceed, we'll subset the data to just involve Adelie and Chinstrap penguins.

```{r penguinsSubset}
penguinsSubset <- subset(penguins, species %in% c("Adelie", "Chinstrap"))
table(penguinsSubset$species)
```

In order to use our machinery for constructing confidence intervals and performing hypothesis tests for the difference in two means, we need some conditions to hold. **What are they?**

1. *Condition 1*
  a. 
  b. 
2. *Condition 2*
  a.
  b.

Let's check these conditions!

*Check of Condition 1*

*Check of Condition 2*

```{r twoSampleNormal, error = T}
qqnorm(penguinsSubset$flipper_length_mm[penguinsSubset$species == "Adelie"],
       main = "Normal Q-Q Plot for Adelie Penguins")
qqline(penguinsSubset$flipper_length_mm[penguinsSubset$species == "Adelie"])

qqnorm(penguinsSubset$flipper_length_mm[penguinsSubset$species == "Chinstrap"],
       main = "Normal Q-Q Plot for Chinstrap Penguins")
qqline(penguinsSubset$flipper_length_mm[penguinsSubset$species == "Chinstrap"])
```

*Are the conditions sufficiently satisfied to proceed?*

Okay, let's perform the t test! We're back to using **formula notation**. Remember how formulas in R work:

> (response variable) ~ (grouping/explanatory variable)

We need to tell R what variable represents the data we're using (the response variable) and which represents the groups. So the first argument to `t.test()` is now the *formula* which describes this information. The rest of the function works the same as before!

```{r twoSampleT}
t.test(flipper_length_mm ~ species,
       data = penguinsSubset,
       mu = 0,
       alternative = "two.sided")
```

Note that the test statistic is $t = -5.6115$, the p-value is $1.297\times 10^{-7}$, and the 95% confidence interval for $\mu_{\text{Adelie}} - \mu_{\text{Chinstrap}}$ is (-7.74, -3.70).

*What's our conclusion?*

## Code Cheat Sheet

The formatting of this will be easier to read in the knitted version of the document.

- **`pt(q, df, lower.tail = TRUE)`**
  - `q` is the x-axis value you want to find an area related to
  - `df` is the degrees of freedom of the $t$ distribution
  - `lower.tail` determines whether `pt()` finds the area to the left or right of `q`. If `lower.tail = TRUE` (the default), it shades to the left. If `lower.tail = FALSE`, it shades to the right.
- **`qt(q, df, lower.tail = TRUE)`**
  - `p` is the probability or area under the curve you want to find an x-axis value for
  - `df` is the degrees of freedom of the $t$ distribution
  - `lower.tail` determines whether `pt()` finds the area to the left or right of `q`. If `lower.tail = TRUE` (the default), it shades to the left. If `lower.tail = FALSE`, it shades to the right.
- **`plotT()`**
  - `df` refers to the degrees of freedom of the distribution to plot. You must provide this value.
  - `shadeValues` is a vector of up to 2 numbers that define the region you want to shade
  - `direction` can be one of `less`, `greater`, `outside`, or `inside`, and controls the direction of shading between `shadeValues`. Must be `less` or `greater` if `shadeValues` has only one element; `outside` or `inside` if two
  - `col.shade` controls the color of the shaded region, defaults to `"cornflowerblue"`
  - `...` lets you specify other graphical parameters to control the appearance of the normal curve (e.g., `lwd`, `lty`, `col`, etc.)
- **`qqnorm(y, ...)`**
  - `y` refers to the variable for which you want to create a Q-Q plot
  - `...` lets you control graphical elements of the plot like `pch`, `col`, etc.
- **`qqline(y, ...)`**
  - `y` refers to the variable for which you created a Q-Q plot
  - `...` lets you control graphical elements of the plot like `pch`, `col`, etc.
  - Function can only be used *after* using `qqnorm()`
- **`t.test(x, alternative = c("two.sided", "less", "greater"), mu = 0, conf.level = 0.95)`**
  - `x` is a vector of data values OR a formula of the form *response ~ group* for two-sample t-tests.
  - `alternative` specifies the direction of the alternative hypothesis; must be one of "two.sided", "less", or "greater"
  - `mu` indicates the true value of the mean (under the null hypothesis); defaults to 0
  - `conf.level` is the confidence level to be used in constructing a confidence interval; must be between 0 and 1, defaults to 0.95


## Try It!
Depending on your instruction method (synchronous remote, asynchronous remote), complete the following exercises collaboratively. 

1. **Asynchronous remote:** Join a Piazza lab group for asynchronous collaboration.
1. **Syncrhonous remote:** In Zoom breakout rooms, work together with your group and use the "Ask for help" button to call your lab instructor.

**Collaborators:** If applicable, replace this text with the names of anyone you collaborated with on this project.

In 2004, the state of North Carolina released a large data set containing information on births recorded in this state. This data set is useful to researchers studying the relation between habits and practices of expectant mothers and the birth of their children. We will work with a random sample of observations from this data set.

```{r readData2}
births <- read.csv("births.csv", stringsAsFactors = T)
```

>**1.** We're going to perform a paired t-test to evaluate the average difference in mother's age and father's age for the population of births in North Carolina in 2004. Clearly state your hypotheses.

$H_0$: replace this text with your null hypothesis

$H_a$: replace this text with your alternative hypothesis

> **2.** Construct a difference variable to use in t-test. It doesn't actually matter what order you do the subtraction in, but for ease of grading, do father's age (`fage`) minus mother's age (`mage`).

```{r tryIt2, error = T}
# Replace this comment with code required for Try It 2. (Remember that this text is a comment, so R ignores it; you can delete it if you want.) If you don't delete it, **start your code on a new line that doesn't start with #**

```

> **3.** Check the conditions necessary to proceed with this test.  

```{r tryIt3, error = T}
# Replace this comment with code required for Try It 3. (Remember that this text is a comment, so R ignores it; you can delete it if you want.) If you don't delete it, **start your code on a new line that doesn't start with #**

```

> **4.** Perform the paired t-test in R and report a conclusion in the context of the data.

```{r tryIt4, error = T}
# Replace this comment with code required for Try It 4. (Remember that this text is a comment, so R ignores it; you can delete it if you want.) If you don't delete it, **start your code on a new line that doesn't start with #**

```

> **5.** Let's now shift to a different question: Do newborns born to mothers who smoke have a different average birthweight than those born to mothers who don't smoke? Set up your hypotheses below.
  
$H_0$: replace this text with your null hypothesis

$H_a$: replace this text with your alternative hypothesis

> **6.** Is this a paired t-test or a test for a difference in two means? Justify your answer.

Replace this text with your answer to Try It 6.

> **7.** Check conditions necessary for performing this hypothesis test. Note that the names of the relevant variables are `weight` and `habit` (for smoking status).

```{r tryIt7, error = T}
# Replace this comment with code required for Try It 7. (Remember that this text is a comment, so R ignores it; you can delete it if you want.) If you don't delete it, **start your code on a new line that doesn't start with #**

```

> **8.** Perform the hypothesis test using `t.test()` and report your conclusion in the context of the research question.

```{r tryIt8, error = T}
# Replace this comment with code required for Try It 8. (Remember that this text is a comment, so R ignores it; you can delete it if you want.) If you don't delete it, **start your code on a new line that doesn't start with #**

```

Replace this text with your written answer for Try It 8.

## Dive Deeper

> **1.** A researcher claims that your results from Try It 8 show that birthweights of babies born to mothers who smoke are *lower* than those born to non-smoking mothers, on average. Without re-running `t.test()`, report whether your results support this claim, providing appropriate numerical support.

Replace this text with your written answer for Dive Deeper 1.


> **2.** Comment on whether or not the following statement is appropriate based on your analysis in Try it 5-8: "maternal smoking causes lower birthweights". If the statement is not appropriate, modify it to make it appropriate.

Replace this text with your written answer for Dive Deeper 2.


> **3.** Comment on whether or not the following statement is appropriate based on your analysis in Try It 1-4: "all fathers in our sample are older than the mothers." If the statement is not appropriate, modify it to make it appropriate.

Replace this text with your written answer for Dive Deeper 3.


> **4.** Determine if the following statement is true or false, and explain your reasoning: "When comparing means of two samples where $n_1 = 20$ and $n_2 = 40$, we can use the normal model for the difference in means since $n_2 \geq 30$. (note $\geq$ gives you the greater-than-or-equal-to sign.)

Replace this text with your written answer for Dive Deeper 4.


## Wrap-Up and Submission

At the top of the document, make sure you've changed the `author` field to your name (in quotes!). If you'd like, change the date as well.

**When you've finished the lab, click the Knit button one last time.**


### Submission instructions
#### If you're using RStudio Cloud
1. In the Files pane, check the box next to `lab12report.html`.
2. Click More > Export...
3. Click Download and save the file on your computer in a folder you'll remember and be able to find later. You can just call the file `lab12report.html` or whatever else you'd like (as long as you remember what you called it).

#### If you're using RStudio installed on your computer
1. locate the `lab12report.html` file on your computer. The file will be saved in the location indicated at the top of the files pane.

#### Submission to Canvas

1. Click the "Assignments" panel on the left side of the page. Scroll to find "Lab 12", and open the assignment. Click "Submit Assignment". 
2. Towards the bottom of the page, you'll be able to choose `lab12report.html` from the folder you saved it in from RStudio Cloud or noted if you're using RStudio Desktop. **You will only be able to upload a .html file -- do not upload any other file type.**
  3. Click "Submit Assignment". You're done! 
